\mythesistitleNOTSYNPOSIS \tnote[fyi]{Synopsis title is ``Generalization of Consecutive Ones Property of Binary Matrices''}

### 2012-01-28 src/020chap2.tex
cleaned up tnotes

\tnote[important]{ADD: have a few lines about organization of chapter}

@@

\begin{definition}{\emph{Graphs that relate to
      COP.\cite[Def.~2.5]{d08phd}}} %
  \label{def:graphwithcop} %
  Let $G$ be a graph and $H$ be a bipartite graph.
  \begin{enumerate}[\ref{def:graphwithcop}--i.]
  \item $G$ is \emph{convex-round} if its adjacency matrix has the
    \CROP.
  \item \label{def::concave-round} $G$ is \emph{concave-round} if its
    augmented adjacency matrix has \CROP. \tnote{cite BHY00}
    \tnote[minor]{add CROP to glossary}
  \item $G$ is an \emph{interval graph} if its vertices can be mapped
    to intervals on the real line \stt two vertices are adjacent \iff
    their corresponding intervals overlap \tnote{cite Ben59, Haj57}.
    $G$ is an interval graph \iff its maximal clique matrix has COP
    \cite{fg65}\footnote{This follows \cite{gh64} which states that the
      maximal cliques of interval graph $G$ can be linearly ordered
      \stt for all $v \in V(G)$, cliques containing $v$ are
      consecutive in the ordering \cite[Th. 8.1]{mcg04}.}
    \begin{enumerate}[a.]
    \item $G$ is a \emph{unit interval graph} if it is an interval
      graph \stt all intervals have the same length.\footnotemark[2]
    \item $G$ is a \emph{proper interval graph} if it is an interval
      graph \stt no interval properly contains another.\footnote{The
        set of unit interval graphs and the set of proper interval
        graphs are the same} \tnote{cite {rob69,gar07} in footnote. pg
        33 dom}
    \end{enumerate}
 \item $G$ is a \emph{circular-arc graph} if its vertices can be
    mapped to a set of arcs on a circle \stt two vertices are adjacent
    \iff their corresponding arcs overlap.\tnote[pressing]{how is
      CO/ROP related?}
  \item $H$ is \emph{convex bipartite on columns (rows)} if its half
    adjacency matrix has \COP on rows (columns).%
    \label{def::convexbi}
  \item $H$ is \emph{biconvex bipartite} or \emph{doubly
      convex}\cite{yc95} if its half adjacency matrix has \COP on both
    rows and columns.
  \item $H$ is \emph{circular convex} if its half adjacency matrix has
    \CROP.
  \end{enumerate}
\end{definition}

Interval graphs\footnote{\cite{mcc04} cites that the problem of
  recognizing interval graphs has significance in molecular
  biology. Interestingly, in the late 1950s, before the structure of
  DNA was well-understood, Seymour Benzer was able to show that the
  intersection graph of a large number of fragments of genetic
  material was an interval graph \cite{sb59}. This was regarded as
  compelling evidence that genetic information was somehow arranged
  inside a structure that had a linear topology which we now know to
  be true from the discovery of linear structure of DNA.}and
circular-arc graphs have a long history in research.  The interest
around them is due to their very desirable property that several
problems that are NP-hard \tnote[minor]{all NPC problems are NPH yes?}
on general graphs, like finding a maximum clique or minimum coloring
or independent set, are polynomial time solvable in these graph
classes \cite{clrs01}.  In a similar fashion, a lot of problems that
are hard on general matrices have efficient solutions on matrices with
\COP or \CROP \cite[more citations pg.\,33]{d08phd}.

@@

It may be noted that one way to design an algorithm to test for \COP
is by deriving one from any interval graph recognition algorithm using
the result {\sc hmpv00}\tnote{cite hmpv00} \tnote{in footnote put the
  theorem 2.7 - dom pg 43} \cite{d08phd} which demonstrates how such a
derivation can be done. However, this does not necessarily yield an
efficient algorithm. We will see results that directly solve the
problem on matrices since it is known that questions~(\ref{q:testcop})
and (\ref{q:copperm}) stated above for \COP are efficiently solvable.
Table~\ref{tab:cophistory} gives a snapshot of these results.

@@

\subsection{Tucker's forbidden submatrices for \COP}
\label{sec:tucker}
% \change{While it is remarkable that the first polynomial time
%   algorithm for \COP testing came in }{The first polynomial time
%   algorithm for \COP testing came in }\cite{fg65} \change{,}{.}
% \change{it is safe to say that the most}{A few years later, a very}
% significant result in understanding \COP came \remove{a few years
%   later} from Tucker which gave a combinatorial (negative)
% characterization of matrices with COP \cite{at72}.

@@

Booth and Lueker in their paper \cite{bl76} gave the first linear
algorithm\footnote{Time complexity is $O\left(m+n+f\right)$ where $m
  \times n$ is the order of the input matrix and $f$ is the number of
  $\un$s in it.} for \cop testing while given a linear time interval
graph recognition algorithm by a simplification of \tnote{cite -
  A. Lempel, S. Even and I. Cederbaum, An Algorithm for Planarity
  Testing of Graphs, Theory of Graphs, ed., P. Rosenstiehl, Gordon and
  Breach, New York, (1967), 215-232.  }'s planarity test algorithm.
% Hsu in wlh01 said BL76 gave a planarity test - "the induced PQ-tree
% algorithm can considerably simplify Booth and Lueker’s modification
% of Lempel, Even and Cederbaum’s planarity test. " 
%
% McConnell in hm03 said the first linear algo for interval graph
% recog was by BL76 - "Benzer’s work and other applications of
% interval graphs motivated a search for eﬃcient algorithms for
% recognizing interval graphs, and for constructing a set of intervals
% to represent the graph when it is one [8]. A linear-time algorithm
% was given in 1976 by Booth and Lueker [2]." 
\cite{bl76} introduces a data structure called \PQtree and their \COP
testing algorithm is a constructive one that outputs a \PQtree if the
input has \COP. A \PQtree represents all the \COP orderings of the
matrix it is associated with. \cite{bl76}'s algorithm uses the fact
that if a matrix has \COP, a \PQtree for it can be constructed. It is
interesting to note that aside from interval graph recognition and
\COP testing, \PQtree is also useful in other applications like
finding planar embeddings of planar graphs \cite{lec67,mcc04} and
recognizing \CROP in a matrix.

@@

Generalized \PQtree or \gPQtree is a data structure defined in
\cite{n89} to represent all orthogonal sets of a set system $\cF$. A
data structure with the same name was later defined in \cite{mcc04} as
part of a {\em substitution decomposition} for a set system $\cF$
\tnote{define substitutio decomp.} and subsequently \cite{mcc04} gives
a new characterization of $\cF$ using a so-called incompatibility
graph. \PQRtree is defined in \cite{mm96} as a data
structure to represent any set system $\cF$ with additional
information in their $R$-nodes if $\cF$ has no \COP.  All three of
these data structures are proposed as generalizations of \cite{bl76}'s
\PQtree and hence produce the \PQtree if $\cF$ has \COP. As a whole,
all these three data strutures are largely identical with their
differences being notional.  In this section, we will discuss the
basic theory that they all hold. We will predominantly use the
terminology from \cite{mm96} and refer to the data structure as
\PQRtree.

@@

\begin{theoremsansproof}[{\cite[Th.~9]{mm96}}] %,\cite[Th.~2.1]{mcc04}
\label{th:pqrfinalmm}
If $\cF$ is a set system with universe $U$ and $|U| \ge 3$ then one of
the following statements hold.
\begin{enumerate}
\item \label{th::it1} $\overline{\cF} = \cT(U)$ 
\item \label{th::it2}$\overline{\cF} = consec(\alpha)$\tnote{what is consec?} for
  some permutation $\alpha$ on $U$ 
\item \label{th::it3}$\overline{\cF} = \power{U}$ 
\end{enumerate}
\end{theoremsansproof}

@@

    &
%     \parbox[t]{\colwidth}
%       {}
    \temptext{PQR tree def. }
    &
    \parbox[t]{\colwidth}
    {
      Decomposition tree of $\cW(\cF)$, $T(\cW(\cF))$
      \temptext{PQR tree def. }\tnote[pressing]{in terms of theory}
    }
    \\\hline

@@

\begin{observation} 
  Consider elements $a, b, c \in U$ and a set system $\cF$ with
  universe $U$. If there is at least one set $S \in \cF$ \stt $a, c
  \in S$ but $b \notin S$, then it is impossible to have a \COP order
  that will place $b$ in between $a$ and $c$. \tnote{also cite the
    inbetweenness problem}
\end{observation}

@@

Thus an edge between two vertices $(a,b), (b,c)$ in $G_I^{\cF}$ means
that there exists no linear order on $U$ that can place $b$ after $a$
and place $c$ after $b$.  For any \COP order, there must not be any
incompatible pairs. Thus it must consist of an independent set $I$ of
the incompatibility graph. Moreover the independent set must have
exactly half the vertex set of $G_I^\cF$ since the \COP order must
involve all elements in $U$. Secondly, the reverse of a \COP order is
also a \COP order. Thus $A_\cF \setminus I$ must also be an
independent set. In other words, $G_I^\cF$ must be bipartite with
partitions being of equal size if $\cF$ has \COP. An odd cycle is a
forbidden structure in a bipartite graph, hence the same is forbidden
in the incompatibility graph of a set system with \COP. In other
words, if the set system has no \COP, its incompatibility graph must
have an odd cycle.  There is also a requirement on the length of this
odd cycle which is formally stated in the following
theorem. \tnote{all refrences to ``following''/abov theorem must be
  chagned to slashrefs. }

@@

\temptext{ So far we have been concerned about matrices that have the
  consecutive ones property. However in real life applications, it is
  rare that data sets represented by binary matrices have COP,
  primarily due to the noisy nature of data available. At the same
  time, COP is not arbitrary and is a desirable property in practical
  data representation \cite{co98,jkckv04,k77}. In this context, there
  are several interesting problems when a matrix does not have COP but
  is ``close'' to having COP or is allowed to be altered to have
  COP. These are the optimization problems related to a matrix which
  does not have COP. Some of the significant problems are surveyed in
  this section.

\tnote{ -- sect 4.1 in cite:d08phd has many results surveyed. hardness
  results, approx. results. results are usually for a class of
  matrices $(a,b)$ where number {\un}s in columns and rows are
  restriced to $a$ and $b$ . -- problem of flipping at most $k$
  entries of $M$ to make it attain COP. this is NP complete
  cite:b75-phd}%
\tnote{(1) scite:lb62 showed that interval graphs are chordal and
  AT-free.} %
\temptext{couple of lines refering to tucker's submatrices. refer
  earlier section.}

Once a matrix has been detected to not have COP (using any of the COT
algorithms mentioned earlier), it is naturally of interest

\begin{enumerate}
\item  to find out
the smallest forbidden substructure (in terms of number of rows and/or
columns and/or number of entries that are {\un}s). \cite{d08phd}
discusses a couple of algorithms which are efficient if the number of
{\un}s in a row is small. This is of significance in the case of
sparse matrices where this number is much lesser than the number of
columns.

\item $(*,\Delta)${\em -matrices} are matrices with no restriction
on number of {\un}s in any column but has at most $\Delta$ {\un}s in
any row. 
{\sc Min COS-R (Min COS-C), Max COS-R (Max COS-C)} are
similar problems which deals with {\em inducing COP} on a matrix.

\begin{enumerate}
\item In the dual problem {\sc Max COS-R (Max COS-C)} the search is for the
maximum number of rows (columns) that induces a submatrix with
COP.
\item In {\sc
  Min COS-R (Min COS-C)} the question is to find the minimum number of
rows (columns) that must be deleted to result in a matrix with COP.
\end{enumerate}

Given a matrix $M$ with no COP, \cite{b75-phd} shows that finding
a submatrix $M'$ with all columns\tnote{check if b75 deals with
  COP col or COP row. also is it any submatrix with k less than r rows
  or submatrix must have all columns?} but a maximum cardinality
subset of rows such that $M'$ has COP is NP complete. \cite{hg02}
corrects an error of the abridged proof of this reduction as given in
\cite{gj79}.  \cite{d08phd} discusses all these problems in detail
giving an extensive survey of the previously existing results which
are almost exhaustively all approximation results and hardness
results. Taking this further, \cite{d08phd} presents new results in
the area of parameterized algorithms for this
problem\tnote{elaborate - what are the results?}.

\item Another problem is to find the minimum number of entries in the matrix
that can be toggled to result in a matrix with COP.  \cite{v85}
discusses approximation of {\sc COP Augmentation} which is the problem
of changing of the minimum number of zero entries to {\un}s so that
the resulting matrix has COP. As mentioned earlier, this problem is
known to be NP complete due to \cite{b75-phd}. \cite{v85} also proves,
using a reduction to the longest path problem, \tnote{or is it a
  survey of another result?  check.} that finding a Tucker's forbidden
submatrix of at least $k$ rows is NP complete. \tnote{how is this
  different from booth's 75 result??}  \tnote{where should this go?
  cite|tz04 (approx submatrix with COP sparse matrices)}

\item \cite{jkckv04} discusses the use of matrices with almost-COP (instead
of one block of consecutive {\un}s, they have $x$ blocks, or {\em
  runs}, of consecutive {\un}s and $x$ is not too large) in the
storage of very large databases.  The problem is that of reordering of
a binary matrix such that the resulting matrix has at most $k$ runs of
{\un}s. This is proved to be NP hard using a reduction from the
Hamiltonian path problem.\tnote{Theorem 2.1 in jkckv} \tnote{(1) A
  connection of COP problem to the travelling salesman problem is also
  introduced. what does this mean? -- COP can be used as a tool to
  reorder $0.5T \le runs(M) \le T$. (2) The optimization version of
  the $k$-run problem, i.e. minimization of number of blocks of ones
  is proven to be NP complete by cite:k77}\tnote{are these two the
  same?} \tnote{what is the reduction?} \tnote{other problems similar
  to COP -- cite:ckl96 (ILP, circ ones, one drop) -- cite:th98
  (generalization of COP - minimax, biotonic column) Tucker}

\end{enumerate}

@@

% \section{Other applications}
% {\tt If any. \tnote{ where should this go?}: (1) cite|jlm97 (application of PQ
%   trees in graphics). (2) helly's theorem citation
%   19XXdgk-Hellystheorem-Danzer-Gruenbaum-Klee}}

%\theendnotes\tnote{remove if none.}

>>>>

\temptext{ADD: have a few lines about organization of chapter}

@@

\begin{definition}{\emph{Graphs that relate to
      COP.\cite[Def.~2.5]{d08phd}}} %
  \label{def:graphwithcop} %
  Let $G$ be a graph and $H$ be a bipartite graph.
  \begin{enumerate}[\ref{def:graphwithcop}--i.]
  \item $G$ is \emph{convex-round} if its adjacency matrix has the
    \CROP.
  \item \label{def::concave-round} $G$ is \emph{concave-round} if its
    augmented adjacency matrix has \CROP. \tnote[pressing]{cite BHY00}
  \item $G$ is an \emph{interval graph} if its vertices can be mapped
    to intervals on the real line \stt two vertices are adjacent \iff
    their corresponding intervals overlap \tnote[pressing]{cite Ben59,
      Haj57}.  $G$ is an interval graph \iff its maximal clique matrix
    has COP \cite{fg65}\footnote{This follows \cite{gh64} which states
      that the maximal cliques of interval graph $G$ can be linearly
      ordered \stt for all $v \in V(G)$, cliques containing $v$ are
      consecutive in the ordering \cite[Th. 8.1]{mcg04}.}
    \begin{enumerate}[a.]
    \item $G$ is a \emph{unit interval graph} if it is an interval
      graph \stt all intervals have the same length.\footnotemark[2]
    \item $G$ is a \emph{proper interval graph} if it is an interval
      graph \stt no interval properly contains another.\footnote{The
        set of unit interval graphs and the set of proper interval
        graphs are the same} \tnote[pressing]{cite {rob69,gar07} in
        footnote. pg 33 dom}
    \end{enumerate}
 \item $G$ is a \emph{circular-arc graph} if its vertices can be
    mapped to a set of arcs on a circle \stt two vertices are adjacent
    \iff their corresponding arcs overlap.\tnote[pressing]{how is
      CO/ROP related?}
  \item $H$ is \emph{convex bipartite on columns (rows)} if its half
    adjacency matrix has \COP on rows (columns).%
    \label{def::convexbi}
  \item $H$ is \emph{biconvex bipartite} or \emph{doubly
      convex}\cite{yc95} if its half adjacency matrix has \COP on both
    rows and columns.
  \item $H$ is \emph{circular convex} if its half adjacency matrix has
    \CROP.
  \end{enumerate}
\end{definition}

Interval graphs\footnote{\cite{mcc04} cites that the problem of
  recognizing interval graphs has significance in molecular
  biology. Interestingly, in the late 1950s, before the structure of
  DNA was well-understood, Seymour Benzer was able to show that the
  intersection graph of a large number of fragments of genetic
  material was an interval graph \cite{sb59}. This was regarded as
  compelling evidence that genetic information was somehow arranged
  inside a structure that had a linear topology which we now know to
  be true from the discovery of linear structure of DNA.}and
circular-arc graphs have a long history in research.  The interest
around them is due to their very desirable property that several
problems that are NP-complete on general graphs, like finding a
maximum clique or minimum coloring or independent set, are polynomial
time solvable in these graph classes \cite{clrs01}.  In a similar
fashion, a lot of problems that are hard on general matrices have
efficient solutions on matrices with \COP or \CROP \cite[more
citations pg.\,33]{d08phd}.

@@

It may be noted that one way to design an algorithm to test for \COP
is by deriving one from any interval graph recognition algorithm using
the result \cite{hmpv00} \cite[Th~2.7]{d08phd} which demonstrates how
such a derivation can be done. However, this does not necessarily
yield an efficient algorithm. We will see results that directly solve
the problem on matrices since it is known that
questions~(\ref{q:testcop}) and (\ref{q:copperm}) stated above for
\COP are efficiently solvable.  Table~\ref{tab:cophistory} gives a
snapshot of these results.

@@

\subsection{Tucker's forbidden submatrices for \COP}
\label{sec:tucker}

@@

Booth and Lueker in their paper \cite{bl76} gave the first linear
algorithm\footnote{Time complexity is $O\left(m+n+f\right)$ where $m
  \times n$ is the order of the input matrix and $f$ is the number of
  $\un$s in it.} for \cop testing while given a linear time interval
graph recognition algorithm by a simplification of \tnote[important]{cite -
  A. Lempel, S. Even and I. Cederbaum, An Algorithm for Planarity
  Testing of Graphs, Theory of Graphs, ed., P. Rosenstiehl, Gordon and
  Breach, New York, (1967), 215-232.  }'s planarity test algorithm.
% Hsu in wlh01 said BL76 gave a planarity test - "the induced PQ-tree
% algorithm can considerably simplify Booth and Lueker’s modification
% of Lempel, Even and Cederbaum’s planarity test. " 
%
% McConnell in hm03 said the first linear algo for interval graph
% recog was by BL76 - "Benzer’s work and other applications of
% interval graphs motivated a search for eﬃcient algorithms for
% recognizing interval graphs, and for constructing a set of intervals
% to represent the graph when it is one [8]. A linear-time algorithm
% was given in 1976 by Booth and Lueker [2]." 
\cite{bl76} introduces a data structure called \PQtree and their \COP
testing algorithm is a constructive one that outputs a \PQtree if the
input has \COP. A \PQtree represents all the \COP orderings of the
matrix it is associated with. \cite{bl76}'s algorithm uses the fact
that if a matrix has \COP, a \PQtree for it can be constructed. It is
interesting to note that aside from interval graph recognition and
\COP testing, \PQtree is also useful in other applications like
finding planar embeddings of planar graphs \cite{lec67,mcc04} and
recognizing \CROP in a matrix.

@@

Generalized \PQtree or \gPQtree is a data structure defined in
\cite{n89} to represent all orthogonal sets of a set system $\cF$. A
data structure with the same name was later defined in \cite{mcc04} as
part of a {\em substitution decomposition} for a set system $\cF$
\tnote[pressing]{define substitutio decomp.} and subsequently \cite{mcc04} gives
a new characterization of $\cF$ using a so-called incompatibility
graph. \PQRtree is defined in \cite{mm96} as a data
structure to represent any set system $\cF$ with additional
information in their $R$-nodes if $\cF$ has no \COP.  All three of
these data structures are proposed as generalizations of \cite{bl76}'s
\PQtree and hence produce the \PQtree if $\cF$ has \COP. As a whole,
all these three data strutures are largely identical with their
differences being notional.  In this section, we will discuss the
basic theory that they all hold. We will predominantly use the
terminology from \cite{mm96} and refer to the data structure as
\PQRtree.

@@

\begin{theoremsansproof}[{\cite[Th.~9]{mm96}}] %,\cite[Th.~2.1]{mcc04}
\label{th:pqrfinalmm}
If $\cF$ is a set system with universe $U$ and $|U| \ge 3$ then one of
the following statements hold.
\begin{enumerate}
\item \label{th::it1} $\overline{\cF} = \cT(U)$ 
\item \label{th::it2}$\overline{\cF} = consec(\alpha)$\tnote[pressing]{what is consec?} for
  some permutation $\alpha$ on $U$ 
\item \label{th::it3}$\overline{\cF} = \power{U}$ 
\end{enumerate}
\end{theoremsansproof}

@@

    &
%     \parbox[t]{\colwidth}
%       {}
    \temptext{PQR tree def. }
    &
    \parbox[t]{\colwidth}
    {
      Decomposition tree of $\cW(\cF)$, $T(\cW(\cF))$
      \temptext{PQR tree def. }
    }
    \\\hline

@@

\begin{observation} 
  Consider elements $a, b, c \in U$ and a set system $\cF$ with
  universe $U$. If there is at least one set $S \in \cF$ \stt $a, c
  \in S$ but $b \notin S$, then it is impossible to have a \COP order
  that will place $b$ in between $a$ and $c$. \tnote[important]{also cite the
    inbetweenness problem}
\end{observation}

@@

Thus an edge between two vertices $(a,b), (b,c)$ in $G_I^{\cF}$ means
that there exists no linear order on $U$ that can place $b$ after $a$
and place $c$ after $b$.  For any \COP order, there must not be any
incompatible pairs. Thus it must consist of an independent set $I$ of
the incompatibility graph. Moreover the independent set must have
exactly half the vertex set of $G_I^\cF$ since the \COP order must
involve all elements in $U$. Secondly, the reverse of a \COP order is
also a \COP order. Thus $A_\cF \setminus I$ must also be an
independent set. In other words, $G_I^\cF$ must be bipartite with
partitions being of equal size if $\cF$ has \COP. An odd cycle is a
forbidden structure in a bipartite graph, hence the same is forbidden
in the incompatibility graph of a set system with \COP. In other
words, if the set system has no \COP, its incompatibility graph must
have an odd cycle.  There is also a requirement on the length of this
odd cycle which is formally stated in the following
theorem. \tnote[pressing]{all refrences to ``following''/abov theorem must be
  chagned to slashrefs. }

@@

\temptext{ -- sect 4.1 in cite:d08phd has many results surveyed. hardness
  results, approx. results. results are usually for a class of
  matrices $(a,b)$ where number {\un}s in columns and rows are
  restriced to $a$ and $b$ . -- problem of flipping at most $k$
  entries of $M$ to make it attain COP. this is NP complete
  cite:b75-phd%
\\(1) scite:lb62 showed that interval graphs are chordal and
  AT-free. %
\\couple of lines refering to tucker's submatrices. refer
  earlier section.

Another problem is to find the minimum number of entries in the matrix
that can be toggled to result in a matrix with COP.  \cite{v85}
discusses approximation of {\sc COP Augmentation} which is the problem
of changing of the minimum number of zero entries to {\un}s so that
the resulting matrix has COP. As mentioned earlier, this problem is
known to be NP complete due to \cite{b75-phd}. \cite{v85} also proves,
using a reduction to the longest path problem, \tnote[pressing]{or is it a
  survey of another result?  check.} that finding a Tucker's forbidden
submatrix of at least $k$ rows is NP complete. \tnote[pressing]{how is this
  different from booth's 75 result??}  \tnote[pressing]{where should this go?
  cite|tz04 (approx submatrix with COP sparse matrices)}

\cite{jkckv04} discusses the use of matrices with almost-COP (instead
of one block of consecutive {\un}s, they have $x$ blocks, or {\em
  runs}, of consecutive {\un}s and $x$ is not too large) in the
storage of very large databases.  The problem is that of reordering of
a binary matrix such that the resulting matrix has at most $k$ runs of
{\un}s. This is proved to be NP hard using a reduction from the
Hamiltonian path problem.\tnote[pressing]{Theorem 2.1 in jkckv}
\tnote[pressing]{(1) A connection of COP problem to the travelling
  salesman problem is also introduced. what does this mean? -- COP can
  be used as a tool to reorder $0.5T \le runs(M) \le T$. (2) The
  optimization version of the $k$-run problem, i.e. minimization of
  number of blocks of ones is proven to be NP complete by
  cite:k77}\tnote{are these two the same?} \tnote{what is the
  reduction?} \tnote{other problems similar to COP -- cite:ckl96 (ILP,
  circ ones, one drop) -- cite:th98 (generalization of COP - minimax,
  biotonic column) Tucker}
}

@@

### 2012-01-28 src/thesis.tex
Did some over all clean up. Added fancy chapter headings. Begun work on opt survey.
* Regular edits. See diff for changes.
* Formatting/layout related changes. stuff i'll forget later.
-------
basically:
- added two side printing
- added headers
- added tag/label markers for draft
- made all chapters/chapter*s start on odd page

technically:
- Added contents to Toc
- increased vglue in ded.tex. adjusted one hspace too. to make them fit in.
- began adding short titles for Toc. (chapter 2)
- added twoside option in LoadClass
- adjusted margins in \RequirePackage{geometry} - reduced left margin.

\usepackage{../lib/mylatexlib}
\usepackage{../lib/mythesislib}
\usepackage{thesisfiguresandtables} % Code for tables are in this
                                % package if it is too long to inline
                                % in .tex file.
%\usepackage[inline] 
%\usepackage[margins] 
\usepackage[finalnew]
         {../lib/trackchanges} % For tracking changes.
                               % :Package options:
                               % [finalold]
                               %   Ignore all of the edits. 
                               %   The document will look as if
                               %   the edits had not been added. 
                               % [finalnew]
                               %   Accept all of the edits. 
                               %   Notes will not be shown in the
                               %   final output. 
                               % [footnotes]
                               %   Added text will be shown
                               %   inline. Removed text and notes
                               %   will be shown as footnotes.  
                               %   This is the default option.
                               % [margins]
                               %   Added text will be shown
                               %   inline. Removed text and notes
                               %   will be shown in the
                               %   margin. Margin notes will be
                               %   aligned with the edits when
                               %   possible. 
                               % [inline]
                               %   All changes and notes will be
                               %   shown inline.

@@

% \usepackage{comment}         % DEBUG ONLY
\includeonly{020chap2}         % DEBUG ONLY 
%\includeonly{030chap3,050appa}         % DEBUG ONLY 
%\includeonly{0031ded}         % DEBUG ONLY

\def \mythesissubmissiondate {13 January 2012}
\def \mythesissubmissionmonth {\MakeUppercase {January 2012}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Certificate
%
\include{001certificate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Acknowledgements
%
\include{002ack}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Dedication?
%
\include{0031ded}%{0021ded}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Abstract
%
\include{003abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table of contents etc.
%
\begin{singlespace}
  \tableofcontents
  \thispagestyle{empty}
  \listoftables \addcontentsline{toc}{chapter}{LIST OF TABLES}
  \listoffigures \addcontentsline{toc}{chapter}{LIST OF FIGURES}
\end{singlespace}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Abbreviations
%
\include{004abbrv}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Notation
%
\include{005notation}
%\pagebreak \clearpage

\pagenumbering{arabic} % The main text will follow from this point so
                       %  set the page numbering to arabic from here on.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Main body Chapters
%
\include{figuresandtables} % Load floats. No rendering.

\include{010chap1}
% turning counter reset. perhaps universal endnote counter is good?
% being schizo about it now. DECIDE! ENDNOTEMARKS DEPEND ON IT.
%\setcounter{endnote}{0} % need to empty the file too. how?! -
                        % FIXED. turns out, just having notes is
                        % enough. or maybe it was fluke.
\include{020chap2}
\include{030chap3}
\include{040chap4}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bibliography.
%
\begin{singlespace}
  \addcontentsline{toc}{chapter}{REFERENCES \tnote[minor]{(i)make names
    in bib file uniform style w.r.t. firstname/initials (ii) have back
  refs - i.e. pages that cite the item.}}

  {\footnotesize%
    \bibliographystyle{alpha}%
    \bibliography{../lib/cop-variants__thesis} %
  }
\end{singlespace}

>>>>

\usepackage{../lib/mylatexlib}
\usepackage{../lib/mythesislib}
\usepackage{thesisfiguresandtables} % Code for tables are in this
                                % package if it is too long to inline
                                % in .tex file.

%% Drafting libraries %%
%
\usepackage[color,notcite,notref]{showkeys}   % To show keys used - useful in
                               % drafts. 
                               % Options: color, notref, notcite,
                               % draft(default), final

\newcommand{\xclearpage}{\cleardoublepage} % cleardoublepage for
                                % twoside. keep empty for one side.
\usepackage[inline] 
%\usepackage[margins] 
%\usepackage[finalnew]
         {../lib/trackchanges} % For tracking changes.
                               % :Package options:
                               % [finalold]
                               %   Ignore all of the edits. 
                               %   The document will look as if
                               %   the edits had not been added. 
                               % [finalnew]
                               %   Accept all of the edits. 
                               %   Notes will not be shown in the
                               %   final output. 
                               % [footnotes]
                               %   Added text will be shown
                               %   inline. Removed text and notes
                               %   will be shown as footnotes.  
                               %   This is the default option.
                               % [margins]
                               %   Added text will be shown
                               %   inline. Removed text and notes
                               %   will be shown in the
                               %   margin. Margin notes will be
                               %   aligned with the edits when
                               %   possible. 
                               % [inline]
                               %   All changes and notes will be
                               %   shown inline.

@@

% \usepackage{comment}         % DEBUG ONLY
%\includeonly{010chap1}         % DEBUG ONLY 
%\includeonly{020chap2}         % DEBUG ONLY 
%\includeonly{030chap3,050appa}         % DEBUG ONLY 
%\includeonly{0031ded}         % DEBUG ONLY

\def \mythesissubmissiondate {31 January 2012}
\def \mythesissubmissionmonth {\MakeUppercase {January 2012}}

%\renewcommand\sectionmark[1]{\markright{\thesection. #1}}
\renewcommand\chaptermark[1]{\markboth{\chaptername\ \thesection. #1}{}}
\renewcommand\sectionmark[1]{\markright{\thesection. #1}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Certificate
%

\xclearpage% \cleardoublepage
\setcounter{page}{1}
\include{001certificate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Acknowledgements
%
\xclearpage
\include{002ack}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Dedication?
%
\xclearpage
\include{0031ded}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Abstract
%
\xclearpage
\include{003abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table of contents etc.
%
\begin{singlespace}
  \xclearpage
  \tableofcontents
  \thispagestyle{empty}
  \xclearpage
  \listoftables %\addcontentsline{toc}{chapter}{LIST OF TABLES}
  \xclearpage
  \listoffigures %\addcontentsline{toc}{chapter}{LIST OF FIGURES}
\end{singlespace}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Abbreviations
%
\xclearpage
\include{004abbrv}

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % Notation
% %
% \xclearpage
% \include{005notation}

\xclearpage
\pagestyle{fancy}%headings}   % For running headers.
\fancyhf{}%            % Clears headers
\fancyhead[LE,RO]{\scriptsize \slshape \rightmark} %{\slshape \rightmark}
\fancyhead[LO,RE]{\scriptsize \slshape \leftmark} %{\slshape \leftmark}
\pagenumbering{arabic} % The main text will follow from this point so
                       %  set the page numbering to arabic from here on.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Main body Chapters
%
\include{010chap1}
% turning counter reset. perhaps universal endnote counter is good?
% being schizo about it now. DECIDE! ENDNOTEMARKS DEPEND ON IT.
%\setcounter{endnote}{0} % need to empty the file too. how?! -
                        % FIXED. turns out, just having notes is
                        % enough. or maybe it was fluke.
\xclearpage
\include{020chap2}
\xclearpage
\include{030chap3}
\xclearpage
\include{040chap4}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bibliography.
%
\begin{singlespace}
%  \addcontentsline{toc}{chapter}{REFERENCES }

  \tnote[minor]{(i)make names
    in bib file uniform style w.r.t. firstname/initials (ii) have back
  refs - i.e. pages that cite the item.}
  {\footnotesize%
    \bibliographystyle{alpha}%
    \bibliography{../lib/cop-variants__thesis} %
  }
\end{singlespace}
